---
title: "Explore AA preferences vs. MAF results"
author: "Gavin Douglas"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output:
  html_document:
    code_folding: hide
    theme: cerulean
    toc: true
    toc_float: true
---


```{r setup}
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(kableExtra))
suppressPackageStartupMessages(library(knitr))
suppressPackageStartupMessages(library(reshape2))
suppressPackageStartupMessages(library(ComplexHeatmap))
library(parallel)
```

```{r read_in}
intab <- read.table('/Users/gavin/Drive/ncsu/aa_selection/aa_selection_zenodo/ecoli_variants/ecoli_seg_subs_w_prefs.tsv.gz',
                    header = TRUE, sep = '\t', stringsAsFactors = FALSE)

intab <- intab[which(intab$N >= 60000), ]

prefs <- colnames(intab)[8:ncol(intab)]

pref_map <- read.table('/Users/gavin/Drive/ncsu/aa_selection/aa_selection_zenodo/aa_metrics/metrics_raw_to_clean.tsv.gz',
                       sep = '\t', header = TRUE, stringsAsFactors = FALSE, row.names = 1)

pref_sd <- sapply(prefs, function(x) { sd(intab[, x]) })
```

# Spearman correlations {.tabset}

Allele frequency vs preference value.

## All mutations

Based on *all* allele frequencies, including singletons.

```{r spearman_w_singletons}
pref_vs_af_p <- numeric()
pref_vs_af_rho <- numeric()
for (pref in prefs) {
  cor_out <- cor.test(intab[, pref], intab$AF, method = 'spearman')
  pref_vs_af_p <- c(pref_vs_af_p, cor_out$p.value)
  pref_vs_af_rho <- c(pref_vs_af_rho, cor_out$estimate)
}
pref_vs_af_spearman <- data.frame(pref = prefs, rho = pref_vs_af_rho, p = pref_vs_af_p)

pref_vs_af_spearman$pref <- pref_map[pref_vs_af_spearman$pref, 'Clean']

# Make a ggplot2 bar graph of the rho values, with the prefs on the y axis.
ggplot(data = pref_vs_af_spearman, aes(x = rho, y = pref)) +
  geom_bar(stat = 'identity') +
  theme_bw() +
  xlab('Spearman correlation coefficient') +
  ylab('Preference') +
  scale_y_discrete(limits = pref_vs_af_spearman$pref[order(pref_vs_af_spearman$rho)])
```

## Excluding singletons


```{r spearman_excluding_singletons}
pref_vs_af_p <- numeric()
pref_vs_af_rho <- numeric()
intab_nosingletons <- intab[which(intab$AC > 1), ]
for (pref in prefs) {
  cor_out <- cor.test(intab_nosingletons[, pref], intab_nosingletons$AF, method = 'spearman')
  pref_vs_af_p <- c(pref_vs_af_p, cor_out$p.value)
  pref_vs_af_rho <- c(pref_vs_af_rho, cor_out$estimate)
}
pref_vs_af_spearman <- data.frame(pref = prefs, rho = pref_vs_af_rho, p = pref_vs_af_p)
pref_vs_af_spearman$pref <- pref_map[pref_vs_af_spearman$pref, 'Clean']

# Make a ggplot2 bar graph of the rho values, with the prefs on the y axis.
ggplot(data = pref_vs_af_spearman, aes(x = rho, y = pref)) +
  geom_bar(stat = 'identity') +
  theme_bw() +
  xlab('Spearman correlation coefficient') +
  ylab('Preference') +
  scale_y_discrete(limits = pref_vs_af_spearman$pref[order(pref_vs_af_spearman$rho)])
```


# Classifying
Classify rare as AC < 10 and frequent as AF > 0.01. Note that singletons were removed.

Note that only sites with at least 60,000 strains non-ambiguously covered are considered here.

```{r define_rare_class}
intab_nosingletons_distinct <- intab_nosingletons
intab_nosingletons_distinct$freq_class <- NA
intab_nosingletons_distinct$freq_class[which(intab_nosingletons_distinct$AC < 10)] <- 'Rare'
intab_nosingletons_distinct$freq_class[which(intab_nosingletons_distinct$AF > 0.01)] <- 'Freq.'
intab_nosingletons_distinct <- intab_nosingletons_distinct[which(! is.na(intab_nosingletons_distinct$freq_class)), ]
```

## Continuous variable differences by freq {.tabset}

### Mean ratio
Ratio of the mean values of substitutions in the rare vs higher groupings.
```{r mean_var_plot}
mean_prefs_raw <- aggregate(. ~ freq_class,
                            data = intab_nosingletons_distinct[, 8:ncol(intab_nosingletons_distinct)],
                            FUN = mean)
mean_prefs_raw <- mean_prefs_raw[, -1]

ratio_df <- data.frame(pref = colnames(mean_prefs_raw),
                       mean_freq_vs_rare = as.numeric(mean_prefs_raw[1, ]) / as.numeric(mean_prefs_raw[2, ]))
ratio_df$pref <- pref_map[ratio_df$pref, 'Clean']

ggplot(data = ratio_df, aes(x = mean_freq_vs_rare, y = pref)) +
  geom_bar(stat = 'identity') +
  theme_bw() +
  xlab('Mean freq. vs rare preference') +
  ylab('Preference') +
  scale_y_discrete(limits = ratio_df$pref[order(ratio_df$mean_freq_vs_rare)])
```

### Cohen's D
```{r cohen_d_plot}

cohend_df <- data.frame(pref = colnames(mean_prefs_raw),
                        cohen_d = (as.numeric(mean_prefs_raw[1, prefs]) - as.numeric(mean_prefs_raw[2, prefs])) / pref_sd)
cohend_df$pref <- pref_map[cohend_df$pref, 'Clean']
ggplot(data = cohend_df, aes(x = cohen_d, y = pref)) +
  geom_bar(stat = 'identity') +
  theme_bw() +
  xlab("Cohen's D") +
  ylab('Preference') +
  scale_y_discrete(limits = cohend_df$pref[order(cohend_df$cohen_d)])
```

# Odd's ratio based on binarizing continuous data {.tabset}

Computed odd's ratio based on a range of cut-offs.

```{r OR_over_range}
ORs <- list()
for (pref_type in prefs) {
  
  if (length(unique(intab_nosingletons_distinct[, pref_type])) >= 100) {
    pref_levels <- seq(from = min(intab_nosingletons_distinct[, pref_type]),
                           to = max(intab_nosingletons_distinct[, pref_type]),
                           length.out = 100)
  } else {
      pref_levels <- sort(unique(intab_nosingletons_distinct[, pref_type]))
  }
      
  # Remove last element (as this doesn't make sense as a lower cut-off).
  pref_levels <- pref_levels[-length(pref_levels)]
    
  var_OR_raw <- mclapply(pref_levels,
                        function(var_cutoff) {
                             cutoff_groups <- ifelse(intab_nosingletons_distinct[, pref_type] <= var_cutoff, 'lower_var', 'higher_var')
                             
                             var_breakdown <- table(cutoff_groups, intab_nosingletons_distinct[, 'freq_class'])
                             
                             var_fisher_out <- fisher.test(var_breakdown)
                             
                             return(data.frame(var=pref_type,
                                               cutoff=var_cutoff,
                                               highervar_higherfreq=var_breakdown[1, 1],
                                               lowervar_higherfreq=var_breakdown[2, 1],
                                               highervar_rarefreq=var_breakdown[1, 2],
                                               lowervar_rarefreq=var_breakdown[2, 2],
                                               OR=var_fisher_out$estimate,
                                               p_value=var_fisher_out$p.value))
                           },
                           mc.cores = 6)
    
    var_OR_combined <- do.call(rbind, var_OR_raw)
    ORs[[pref_type]] <- var_OR_combined
}

OR_var <- do.call(rbind, ORs)
```

```{r continuous_OR_full}
OR_var$var <- factor(OR_var$var, levels = prefs)

OR_var$Sig <- ifelse(OR_var$p_value < 0.05, 'yes', 'no')

OR_var <- OR_var[which(! is.infinite(OR_var$OR)), ]

ggplot(data = OR_var, aes(x = cutoff, y = OR, colour = Sig)) +
    geom_point() +
    scale_colour_manual(values=c('grey80', 'black')) +
    theme_bw() +
    facet_wrap(. ~ var, scales = 'free_x') +
    ylab("Odd's ratio (rare vs higher MAF)") +
    xlab('Cutoff (for calling high and low values for variable)') +
    theme(plot.title = element_text(hjust = 0.5))
```

# ROC curve based on binarizing continuous data {.tabset}
Get true positive and false positive rate estimates per test. Here, a 'positive' is a higher frequency allele.
This is based on the (questionable) assumption that values below the cut-off are expected to be "rare" and those above are expected to be "higher" frequency.

```{r compute_ROC_values}
OR_var$positives <- OR_var$lowervar_higherfreq + OR_var$highervar_higherfreq
OR_var$negatives <- OR_var$lowervar_rarefreq + OR_var$highervar_rarefreq

OR_var$TPR <- OR_var$highervar_higherfreq / OR_var$positives
OR_var$FPR <- OR_var$highervar_rarefreq / OR_var$negatives
```

```{r continuous_ROC_full, fig.width=9, fig.height=5}
OR_var$var <- as.character(OR_var$var)
OR_var$var <- pref_map[OR_var$var, 'Clean']

OR_var <- OR_var[order(OR_var$TPR, OR_var$FPR), ]

AUC_tab <- data.frame(var = sort(unique(OR_var$var)), AUC = NA)
rownames(AUC_tab) <- AUC_tab$var
for (var in rownames(AUC_tab)) {
  var_subset <- OR_var[OR_var$var == var, ]
  AUC_tab[var, 'AUC'] <- sum(diff(var_subset$FPR) * (var_subset$TPR[-1] + var_subset$TPR[-length(var_subset$TPR)]) / 2)
}

OR_var_plot <- ggplot(data = OR_var, aes(x = FPR, y = TPR, colour = var)) +
  geom_line() +
  theme_bw() +
  geom_abline(intercept = 0, slope = 1, linetype = 2, color = 'grey') +
  ylab("True Positive Rate") +
  xlab("False Positive Rate") +
  theme(plot.title = element_text(hjust = 0.5))

AUC_barplot <- ggplot(data = AUC_tab, aes(y = var, x = AUC)) +
  geom_bar(stat = 'identity', fill = 'steelblue') +
  geom_text(aes(label = round(AUC, 3)), vjust = -0.5, hjust = 1.5, col = "white") +
  theme_bw() +
  xlab('AUC') +
  ylab('Variable') +
  theme(plot.title = element_text(hjust = 0.5)) +
  scale_y_discrete(limits = AUC_tab$var[order(AUC_tab$AUC)])

AUC_barplot
```


# Prop of most extreme values rare

Rather than classifying rare and frequent, which doesn't make any sense realistically,
what does make (more) sense is being able to classify substitutions with extremely low
probabilities as rare (or invariant, as can be done in future analyses...).

Look at proportion of substitutions that are rare based on the bottom 5% of the scores for each preference.
A higher proportion here is better!

```{r prop_extreme_rare}
raw <- list()

for (cutoff in c(0.001, 0.01, 0.1)) {

  prop_rare <- c()

  for (pref in prefs) {
    pref_quantile <- quantile(intab_nosingletons_distinct[, pref], probs = cutoff)
    tab_subset <- intab_nosingletons_distinct[which(intab_nosingletons_distinct[, pref] <= pref_quantile), ]
    prop_rare <- c(prop_rare, length(which(tab_subset$freq_class == 'Rare')) / nrow(tab_subset))
  }
  
  prep_prop_rare_at_extreme <- data.frame(cutoff=as.character(cutoff), pref = prefs, prop_rare = prop_rare)
  prep_prop_rare_at_extreme[order(prep_prop_rare_at_extreme$prop_rare, decreasing = TRUE), ]
  
  prep_prop_rare_at_extreme$pref <- pref_map[prep_prop_rare_at_extreme$pref, 'Clean']
  
  raw[[as.character(cutoff)]] <- prep_prop_rare_at_extreme
  
}

combined_prop_rare <- do.call(rbind, raw)

combined_prop_rare$pref <- as.character(combined_prop_rare$pref)

overall_rare_prop <- length(which(intab_nosingletons_distinct$freq_class == 'Rare')) / nrow(intab_nosingletons_distinct)


combined_prop_rare$cutoff <- factor(combined_prop_rare$cutoff, levels = c('0.1', '0.01', '0.001'))

# Only keep the original Grantham and Miyata as it is information overflow otherwise...
prefs_to_rm <- c("Grantham (update, recalc.)", "Grantham (update, min-max)", "Grantham (recalc.)", "Grantham (min-max)",
                 "Miyata (recalc.)", "Miyata (min-max)", "Miyata (update, min-max)", "Miyata (update)")

combined_prop_rare <- combined_prop_rare[which(! combined_prop_rare$pref %in% prefs_to_rm), ]
combined_prop_rare$pref[which(combined_prop_rare$pref == "Grantham (orig.)")] <- 'Grantham'
combined_prop_rare$pref[which(combined_prop_rare$pref == "Miyata (orig.)")] <- 'Miyata'

mean_combined_prop_rare <- aggregate(prop_rare ~ pref, data = combined_prop_rare, FUN = mean)


ggplot(data = combined_prop_rare, aes(x = prop_rare, y = pref, fill = cutoff)) +
    geom_bar(stat = 'identity', position = 'dodge') +
    theme_bw() +
    xlab('Proportion of subs. in cut-off proportion\nof lowest preferences that are rare') +
    ylab('Preference') +
    scale_y_discrete(limits = mean_combined_prop_rare$pref[order(mean_combined_prop_rare$prop_rare)]) +
  scale_fill_manual(values = c('0.001' = 'grey30', '0.01' = 'grey60', '0.1' = 'grey90')) +
  coord_cartesian(xlim = c(0.9, 1)) +
  labs(fill = "Cut-off") +
    geom_vline(xintercept = overall_rare_prop, linetype = 'dashed', color = 'blue', size = 0.5) +
  guides(fill = guide_legend(reverse = TRUE))
```

# Session info {.tabset}

## Hide

## Show

```{r show_info}
sessionInfo()
```
